A hidden Active Galactic Nuclei population: the first radio luminosity functions constructed by physical process instead of galaxy classification

### 2024-10-01 src/tex/ms.tex
[showyourwork] overleaf sync

% Author list
\author[L.K. Morabito]{\parbox{\textwidth}{Leah K. Morabito$^{1,2}$\thanks{E-mail: leah.k.morabito@durham.ac.uk}\orcidlink{0000-0003-0487-6651}
J.M.G.H.J. de Jong$^{3}$,
F. Sweijen$^{1}$,
R. Kondapally$^{4}$,
P.N. Best$^{4}$,
B. Yue$^{4}$, and friends
\\}\\ 
$^{1}$Centre for Extragalactic Astronomy, Department of Physics, Durham University, Durham DH1 3LE, UK \\
$^{2}$Institute for Computational Cosmology, Department of Physics, Durham University, South Road, Durham DH1 3LE, UK \\ 
$^{3}$Leiden \\
$^{4}$Edinburgh \\}

@@

% Abstract with filler text
\begin{abstract}
Both star formation (SF) and Active galactic nuclei (AGN) play an important role in galaxy evolution. Quantifying their relative importance in a statistical way can be done using radio luminosity functions. Until now these have relied on galaxy classifications, with sources which have a mixture of radio emission from SF and AGN assigned to either a star-forming galaxy or an AGN. This can lead to the under-estimation of the relevance of AGN, or an over-estimation of SF. Brightness temperature methods were demonstrated to be effective at 144 MHz using sub-arcsecond imaging with the International LOFAR Telescope (ILT) in \cite{morabito_identifying_2022}. We use the combination of sub-arcsecond and arcsecond resolution imaging of 7,497 sources in the Lockman Hole and ELAIS-N1 fields to identify AGN components in the high resolution images and subtract them from the total flux density, leaving behind flux density from SF only. We construct, for the first time, radio luminosity functions by physical process rather than galaxy classification, revealing a hidden AGN population at \llof $<10^{24}$\wphz . This population is $\sim$50 percent more than expected for $0.5<z<2.0$. The star forming population is less impacted, with only $\sim$10 - 20 percent less star formation than expected. The contribution of `hidden' AGN to the overall radio population, especially at \llof $<10^{24}$\wphz , is significant, and only revealed thanks to the combination of the ILT's resolution and wide field of view. \vspace{3.7in}
\end{abstract}

% Main body with filler text
\section{Introduction}
\label{sec:intro}
The extent to which active galactic nuclei (AGN) impact galaxy formation is a major open question in astrophysics. It is clear that there is an interaction, from both observational and theoretical standpoints. Observations have revealed tight scaling relations between the mass of a super-massive black hole and its host galaxy \citep[see, e.g.][and references therein]{kormendy_coevolution_2013}, while cosmological simulations require some form of AGN feedback \citep{bower_breaking_2006,croton_many_2006} to be able to reproduce the observed galaxy population. 
%Galaxies build up their stellar mass through star formation, while super-massive black holes are expected to provide either negative or positive feedback to either suppress or stimulate galaxy growth, although whether we can definitively pinpoint which type of feedback is an open question \citep{ward_cosmological_2022}. 
The cosmic history of AGN activity and star formation (SF), and by extension the interplay between them, are interlinked. It is necessary to quantify the contribution of each process to understand the importance of either or both of them. Relying on overall galaxy classification can over- or under-estimate the contributions of each process; for example, by considering all of the radio emission in a galaxy classified as `star-forming' to be due to star formation, when some may be due to AGN. Quantifying this is the goal of this letter.

Decomposing AGN activity from star formation is a difficult problem at any wavelength. Efforts have been made to do this via mid-infrared (MIR) spectral decomposition using template fitting \citep[e.g.,][]{laurent_mid-infrared_2000,hernan-caballero_resolving_2015,li_active_2024}. Multi-wavelength approaches use spectral energy distribution (SED) fitting, which utilises information from a much broader number of observed bands to jointly fit galaxy and AGN components \citep[e.g.][]{calistro_rivera_agnfitter_2016,boquien_cigale_2019,pacifici_art_2022}. However, SED fitting is costly as it relies on observations using many different instruments over the same part of the sky, which then have to be carefully cross-matched. Varying selection effects across the different bands can also present a problem.

An alternative method to unambiguously identify the AGN component(s) in a galaxy is through brightness temperature, $T_b$, measurements in the radio waveband. Radio emission has the advantage of being unobscured by dust or gas, meaning there are minimal selection effects. %Brightness temperature is a non-physical temperature which is defined as the temperature of a blackbody which would produce the observed surface brightness. 
Models of radio emission from SF predict an upper limit to the surface brightness that can be produced, even in starburst galaxies \citep{condon_radio_1992}. Surface brightness which is detected above this limit therefore has to be attributed to AGN activity. Brightness temperature is inversely proportional to frequency and resolution, $T_b \propto \nu^{-2}\Omega^{-1}$ (where $\Omega \propto \theta_{maj}\theta_{min}$). Traditionally brightness temperature measurements require Very Long Baseline Interferometry (VLBI) which achieve milli-arcsec resolutions at GHz frequencies \citep[e.g.,][]{middelberg_mosaiced_2013,herrera_ruiz_faint_2017,radcliffe_nowhere_2018} to achieve enough sensitivity to make meaningful measurements. In \cite{morabito_identifying_2022} (hereafter, \citetalias{morabito_identifying_2022}) we demonstrated using sub-arcsecond resolution (0.\sarc\ 3) at 144 MHz with the LOw Frequency ARray \citep[LOFAR;][]{van_haarlem_lofar:_2013}. The advantage of LOFAR is its wide field of view: $\sim$6 deg$^2$ from a single observation versus a few to a couple hundred arcmin$^2$ with VLBI at GHz frequencies, depending on whether one is imaging at the phase centre of the observation or using multi-phase centres for correlation \citep{deller_difx-2:_2011,morgan_vlbi_2011}. For example, using an 8 hour LOFAR observation of the Lockman Hole \citep{sweijen_deep_2022} we identified 940 AGN via brightness temperature \citepalias{morabito_identifying_2022}, providing a sample two orders of magnitude larger than any VLBI sample of AGN identified at GHz frequencies.

As part of the LOFAR Two-metre Sky Survey Deep Fields \citep{sabater_lofar_2021,tasse_lofar_2021} campaign, we have now doubled the areal coverage of 0.\sarc\ 3 resolution wide-field imaging. \cite{de_jong_into_2024} recently published their 0.\sarc\ 3 resolution image of the ELAIS-N1 field, which reaches a depth of 14 $\mu$Jy$\,$bm$^{-1}$ using 32 hours of observations. They also demonstrate the flexible resolution which can be achieved with LOFAR, by imaging at 0.\sarc\ 6 and 1.\sarc\ 2 resolution, which are both complementary to the existing 6\sarc\ \ Deep Fields Data Release 1 (DR1). Again this has an advantage over VLBI with instruments at GHz frequencies, which require either observing at different frequency bands or with other instruments to acquire information on the flux density of sources at other scales. As described in \S~5 of \citepalias{morabito_identifying_2022}, for sources which are unresolved in the 0.\sarc\ 3 resolution images and do not exhibit a radio excess over that predicted the star formation rate (SFR) derived from SED fitting \citep[see][for more details]{best_lofar_2023}, we can separate the AGN activity from SF using a combination of the 0.\sarc\ 3 and 6\sarc\ \ resolution images.

The 2,483 sources in Lockman Hole \citep{sweijen_deep_2022} and now the 13,058 sources in ELAIS-N1 \citep{de_jong_into_2024}, represents a massive step towards robust statistical studies where the population can be broken into bins of redshift and stellar mass to investigate the AGN and SFG populations in detail. \cite{cochrane_lofar_2023} and \cite{kondapally_cosmic_2022} constructed radio luminosity functions (RLFs) from the 6\sarc\ \ resolution imaging that formed the Deep Fields DR1 for the star forming galaxies and radio-excess AGN, respectively. Here we construct RLFs, for the first time, by process rather than overall galaxy classification. We do this by using a combination of the 0.\sarc\ 3 and 6\sarc\ \ resolution imaging from Lockman Hole and ELAIS-N1. We investigate the cosmic evolution of AGN activity and SF by dividing the sample into redshift bins.

In this paper, we first describe the data and methods in \S~\ref{sec:data}, followed by an overview of the methods in \S~\ref{sec:methods}. Section~\ref{sec:results} describes the results, followed by 
% discussion and 
conclusions in 
%\S~\ref{sec:discussion} and
\S~\ref{sec:conclusions}
%, respectively
.

Throughout this paper, we use a $H_0 = 70$ km$\,$s$^{-1}\,$Mpc$^{-1}$, $\Omega_M = 0.3$, and $\Omega_{\Lambda}= 0.7$ cosmology. Radio flux density is defined as $S\propto \nu^{\alpha}$ where $\nu$ is frequnecy and $\alpha$ is radio spectral index.

%Stuff and things \citep{morabito_sub-arcsecond_2022}.

@@

\subsection{LoTSS Deep Fields DR1}
\label{subsec:dr1}
The LOFAR Two-metre Sky Survey (LoTSS) has both a wide-area \citep{shimwell_lofar_2017,shimwell_lofar_2021,shimwell_lofar_2022} and Deep Fields \citep{tasse_lofar_2021,sabater_lofar_2021} component. The Deep Fields Data Release 1 (DR1) in 2021 included Boötes, Lockman Hole, and ELAI-N1, reaching sensitivities of 30, 23, and 20 \muJybm , using $\sim$80, $\sim$100, and $\sim$160 hours, respectively. The resolution of these images is all 6\sarc . Alongside these deep radio images and their associated catalogues, \cite{kondapally_lofar_2021} provided a careful compilation of the ancillary deep data at other wavebands (from far-infrared to ultraviolet) and cross-matching to the radio catalogues, followed by photometric redshifts and stellar mass estimates in \cite{duncan_lofar_2021}. Collectively these catalogues provide an excellent resource, and we refer the reader for more details to the individual papers cited above.

Based on the cross-matched multi-wavelength catalogue \cite{kondapally_lofar_2021}, \cite{best_lofar_2023} (hereafter, \citetalias{best_lofar_2023}) carried out detailed SED fitting for all radio-detected sources across the three deep fields, using the multi-wavelength data. Full details can be found in their paper, but here we outline the relevant details. All sources detected above 5$\sigma$ in their respective radio images had their far-infrared (FIR) to ultraviolet (UV) data fit using four different SED fitting software packages. These were: MAGPHYS \citep{da_cunha_simple_2008}, BAGPIPES \citep{carnall_inferring_2018,carnall_how_2019}, CIGALE \citep{burgarella_star_2005,noll_analysis_2009,boquien_cigale_2019}, and AGN\textsc{fitter} \citep{calistro_rivera_agnfitter_2016}. The first two (MAGPHYS and BAGPIPES) use an energy balance approach, where optical to UV energy absorbed by dust must be transferred to thermal emission in the FIR to sub-mm.  CIGALE also uses the same energy-balance approach, but includes an AGN component. The final code, AGN\textsc{fitter} is designed to work in cases where the UV light and IR emission are spatially distinct, and thus the energy balance assumption does not hold. Many AGN fall under this umbrella. Each galaxy was fit with all four codes (twice for CIGALE, using different AGN components) and a `consensus' value found for the derived galaxy properties using information from the goodness-of-fits.

The derived properties used in this paper are the consensus SFRs, and the galaxy classifications which arise from the SED fitting. \citetalias{best_lofar_2023} use the relationship between radio luminosity and SFR to define sources as either \textit{radio excess} or not, based on the expected relationship between radio luminosity and SFR. The AGN components used to fit the multi-wavelength information indicate whether a source is a radiatively efficient AGN (i.e., is there evidence for a disk/torus structure) or not. Combining these provides the following classes:
\begin{itemize}
    \item Radio excess + radiatively efficient (\textit{`radio-loud'})
    \item Radio excess + not radiatively efficient (\textit{`radio-loud'})
    \item Radiatively efficient + not radio excess (\textit{`radio-quiet'})
    \item Radiatively inefficient + not radio excess
\end{itemize}
The first three classes are AGN, and the final class is star forming galaxies (SFGs).  Five percent of all sources are Unclassified, as they do not have a radio excess and the fits could not determine between radiatively efficient or inefficient.

@@

Data for LoTSS is recorded with all available LOFAR stations, which are spread across nine different European countries, with a preponderance of stations in the Netherlands. The longest baseline (currently Ireland to Latvia) is $\sim$2,000 km, providing unprecedented resolution at MHz frequencies. While LoTSS uses only the stations in the Netherlands for its standard data processing, post-processing of LoTSS data to achieve sub-arcsecond resolution is ongoing both for individual sources with $S>10\,$mJy in the wide-area survey \citep{morabito_sub-arcsecond_2022} and for the entire field of view in the Deep Fields \citep[][Escott et al., in prep; Bondi et al., in prep]{sweijen_deep_2022,de_jong_into_2024}. %While the full field of view imaging is currently too computationally expensive for the wide area, we are focusing on the Deep Fields. 
Due to the fact that the station layout of international stations is larger than that of stations in the Netherlands, the station beams have a smaller field of view. Coupled with bandwidth and time smearing, this limits the field of view effectively to a radius of 1.24 deg, which still provides 6.25 deg$^2$ of coverage\footnote{The imaged field of view is square with 2.5 deg per side.}.

Both Lockman Hole and ELAIS-N1 have been imaged at 0.\sarc\ 3 resolution \citep[][respectively]{sweijen_deep_2022,de_jong_into_2024}. The resulting images have median rms noise of 34 \muJybm\ (Lockman Hole) and 14 \muJybm\ (ELAIS-N1), although the noise is lowest in the centre and increases radially in both images. Catalogues from these images were processed to group individual Gaussian islands into complete sources, and remove duplicates. The resulting catalogues contain 2,316 and 13,058 sources above 5$\sigma$, for Lockman Hole and ELAIS-N1 respectively. Although \cite{de_jong_into_2024} also imaged ELAIS-N1 at 0.\sarc\ 6 and 1.\sarc\ 2, for consistency with Lockman Hole we use only the 0.\sarc\ 3 and 6\sarc\ resolution images.

@@

Thanks to the added collecting area of the international stations, which increases the sensitivity of the array, the 0.\sarc\ 3 resolution image for ELAIS-N1 is \textit{deeper} than the Deep Fields DR1. To ensure consistently built samples, and make use of DR1 information, we limit the high-resolution catalogues. For both fields, 
%The derived galaxy properties from \citetalias{best_lofar_2023} (see \S~\ref{subsec:dr1}) were for LoTSS Deep Fields DR1, and thus we have to limit the source catalogues we wish to use. For both Lockman Hole and ELAIS-N1 catalogues derived from the 0.\sarc\ 3 resolution images, 
we first find the local rms in the companion 6\sarc\ \ resolution noise maps\footnote{\label{dr1}Available at \href{https://lofar-surveys.org/deepfields.html}{https://lofar-surveys.org/deepfields.html}} at the position of each source in the 0.\sarc\ 3 resolution catalogue. Using the peak brightness of the source in the catalogue from the 0.\sarc\ 3 resolution image, we calculate the peak-to-noise ratio and only select sources which would be detectable at the 5$\sigma$ limit in the 6\sarc\ resolution image, effectively ensuring all sources will have a match in DR1. We then cross-match based on radio position, to provide the radio information from the catalogues derived from both the 0.\sarc\ 3 resolution image and the 6\sarc\ \ resolution image. We trim the area down for each field using their respective optical starmasks\textsuperscript{\ref{dr1}} by selecting sources which are not contained in the null areas of these masks. This removes sources where the multi-wavelength information is impacted by bright stars. Finally, we trim the area down to the 2.5 deg$\times$2.5 deg area which is covered by the 0.\sarc\ 3 resolution image. We are left with \variable{output/lockman_detectable.txt}sources in Lockman Hole, and \variable{output/en1_detectable.txt}sources in ELAIS-N1. We use the source name from the catalogue derived from the 6\sarc\ \ resolution to cross-match to the catalogue from \citetalias{best_lofar_2023}.

We determine whether sources are resolved or unresolved in the same manner as \S~3.1 of \cite{shimwell_lofar_2019,shimwell_lofar_2022}. Truly unresolved sources in radio maps should have equal peak brightness and integrated flux density, but LOFAR maps are impacted by smearing due to the ionosphere. This causes resolved sources to have a lower peak brightness when compared to the integrated flux density, and they can appear spatially extended. We start by selecting sources which are fit by a single Guassian component ({\ttfamily S\_CODE = `S'}), and make a cut to remove sources which are larger than a multiple of the beam size. For the Lockman Hole, we use a factor of 3, but for ELAIS-N1 we find that a factor of 4 provided a better sample. This may be because half of the data for ELAIS-N1 had been pre-averaged to 2 sec integration times, which will increase the time smearing \citep[see \S~2 in][]{de_jong_into_2024}. We fit a Sigmoid function to find the upper 99.9th percentile of sources considering their integrated flux density to peak brightness ratio as a function of the local signal to noise. Anything above this boundary is considered to be resolved. We also treat sources which do not have {\ttfamily S\_CODE = `S'} as resolved.

@@

We follow the same method of constructing RLFs as in \cite{kondapally_cosmic_2022} and \cite{cochrane_lofar_2023}. The 0.\sarc\ 3 resolution image is smaller than the 6\sarc\ resolution image, so we first reproduce the published RLFs (Fig.~\ref{fig:rlfs}, left panel) using the galaxy classifications from \citetalias{best_lofar_2023} to check that we have implemented the method appropriately, and that the smaller area yields the same results. We use the completeness corrections from the respective papers. The only correction we do not implement is the photo-$z$ correction used in \cite{cochrane_lofar_2023}. The authors note that this correction is negligible, and as it is not implemented in \cite{kondapally_cosmic_2022} we chose to keep the method for RLF construction in this paper consistent across both samples. We use the same redshift range, $0.003 < z < 0.3$, to verify that we have good agreement with the published RLFs.

>>>>

% Author list
\author[L.K. Morabito]{\parbox{\textwidth}{Leah K. Morabito$^{1,2}$\thanks{E-mail: leah.k.morabito@durham.ac.uk}\orcidlink{0000-0003-0487-6651}
J.M.G.H.J. de Jong$^{3}$,
F. Sweijen$^{1}$,
R. Kondapally$^{4}$,
P.N. Best$^{4}$,
B.-H. Yue$^{4}$\orcidlink{0009-0009-8935-2929}, 
T.W. Shimwell$^{3,5}$, and friends
\\}\\ 
$^{1}$Centre for Extragalactic Astronomy, Department of Physics, Durham University, Durham DH1 3LE, UK \\
$^{2}$Institute for Computational Cosmology, Department of Physics, Durham University, South Road, Durham DH1 3LE, UK \\ 
$^{3}$Leiden Observatory, Leiden University, PO Box 9513, 2300 RA Leiden, The Netherlands \\
$^{4}$Institute for Astronomy, University of Edinburgh Royal Observatory, Blackford Hill, Edinburgh, EH9 3HJ, UK \\
$^{5}$ASTRON, Netherlands Institute for Radio Astronomy, Oude Hoogeveensedijk 4, 7991 PD, Dwingeloo, the Netherlands \\}

@@

% Abstract with filler text
\begin{abstract}
Both star formation (SF) and Active Galactic Nuclei (AGN) play an important role in galaxy evolution. Quantifying their relative importance in a statistical way can be done using radio luminosity functions. Until now these have relied on galaxy classifications, with sources which have a mixture of radio emission from SF and AGN assigned to either a star-forming galaxy or an AGN. This can lead to the under- or over-estimation of the relevance of AGN. Brightness temperature methods were demonstrated to be effective at separating radio emission from AGN and SF at 144 MHz using sub-arcsecond imaging with the International LOFAR Telescope (ILT). We use the combination of sub-arcsecond and arcsecond resolution imaging of 7,497 sources in the Lockman Hole and ELAIS-N1 fields to identify AGN components in the high resolution images and subtract them from the total flux density, leaving behind flux density from SF only. We construct, for the first time, radio luminosity functions by physical process rather than galaxy classification, revealing a hidden AGN population at \llof $<10^{24}$\wphz . This population is $\sim$50 percent more than expected for $0.5<z<2.0$ when comparing to RLFs by galaxy classification. The star forming population is less impacted, with only $\sim$10 - 20 percent less star formation than expected. The contribution of `hidden' AGN to the overall radio population, especially at \llof $<10^{24}$\wphz\ can have significant implications for the cosmic star formation rate and kinetic luminosity densities. This is only revealed thanks to the combination of the ILT's resolution and wide field of view. \vspace{3.7in}
\end{abstract}

% Main body with filler text
\section{Introduction}
\label{sec:intro}
The extent to which active galactic nuclei (AGN) impact galaxy formation is a major open question in astrophysics. It is clear that they influence each other, from both observational and theoretical standpoints. Observations have revealed tight scaling relations between the mass of a super-massive black hole and its host galaxy \citep[see, e.g.][and references therein]{kormendy_coevolution_2013}, while cosmological simulations require some form of AGN feedback \citep{bower_breaking_2006,croton_many_2006} to be able to reproduce the observed galaxy population. 
%Galaxies build up their stellar mass through star formation, while super-massive black holes are expected to provide either negative or positive feedback to either suppress or stimulate galaxy growth, although whether we can definitively pinpoint which type of feedback is an open question \citep{ward_cosmological_2022}. 
The cosmic histories of AGN activity and star formation (SF), and by extension the interplay between them, are interlinked. It is necessary to quantify the contribution of each process to understand the importance of either or both of them. Relying on overall galaxy classification can over- or under-estimate the contributions of each process; for example, by considering all of the radio emission in a galaxy classified as `star-forming' to be due to star formation, when some may be due to AGN. Quantifying these contributions is the goal of this letter.

Decomposing AGN activity from star formation (SF) on a per-galaxy basis is a difficult problem at any wavelength. Efforts have been made to do this via mid-infrared (MIR) spectral decomposition using template fitting \citep[e.g.,][]{laurent_mid-infrared_2000,hernan-caballero_resolving_2015,li_active_2024}. Multi-wavelength approaches use spectral energy distribution (SED) fitting, which uses information from a much broader number of observed bands to jointly fit galaxy and AGN components \citep[e.g.][]{calistro_rivera_agnfitter_2016,boquien_cigale_2019,pacifici_art_2023}. However, SED fitting is costly as it relies on observations using many different instruments over the same part of the sky, which then have to be carefully cross-matched, and there can be degeneracies in the fits. Varying selection effects across the different bands can also present a problem.

An alternative method to unambiguously identify the AGN component(s) in a galaxy is through brightness temperature, $T_b$, measurements in the radio waveband. Observations of radio emission have the advantage of being unobscured by dust or gas, so there are minimal observational biases. %Brightness temperature is a non-physical temperature which is defined as the temperature of a blackbody which would produce the observed surface brightness. 
Models of radio emission from SF predict an upper limit to the surface brightness that can be produced, even in starburst galaxies \citep{condon_radio_1992}. Surface brightness which is detected above this limit therefore has to be attributed to AGN activity. Brightness temperature is inversely proportional to frequency and resolution, $T_b \propto \nu^{-2}\Omega^{-1}$ (where $\Omega \propto \theta_{\text{maj}}\theta_{\text{min}}$). Traditionally brightness temperature measurements require Very Long Baseline Interferometry (VLBI) which reach milli-arcsec resolutions at GHz frequencies \citep[e.g.,][]{middelberg_mosaiced_2013,herrera_ruiz_faint_2017,radcliffe_nowhere_2018} to achieve enough sensitivity to make meaningful measurements. In \cite[][hereafter, \citetalias{morabito_identifying_2022}]{morabito_identifying_2022} we demonstrated that this can be done using sub-arcsecond resolution (0.\sarc\ 3) at 144 MHz with the LOw Frequency ARray \citep[LOFAR;][]{van_haarlem_lofar:_2013}. The advantage of LOFAR is its wide field of view: $\sim$6 deg$^2$ from a single observation versus a few to a couple hundred arcmin$^2$ with VLBI at GHz frequencies, depending on whether one images a single phase centre or uses multi-phase centres for correlation \citep{deller_difx-2:_2011,morgan_vlbi_2011}. For example, using an 8 hour LOFAR observation of the Lockman Hole \citep{sweijen_deep_2022} we identified 940 AGN via brightness temperature \citepalias{morabito_identifying_2022}, providing a sample two orders of magnitude larger than any VLBI sample of AGN identified at GHz frequencies, using a fraction of the observing time.

As part of the LOFAR Two-metre Sky Survey Deep Fields campaign \cite[][hereafter, \citetalias{best_lofar_2023}]{best_lofar_2023}, we have now doubled the areal coverage of 0.\sarc\ 3 resolution wide-field imaging. \cite{de_jong_into_2024} recently published their 0.\sarc\ 3 resolution image of the ELAIS-N1 field, which reaches a depth of 14 $\mu$Jy$\,$bm$^{-1}$ using 32 hours of observations. They also demonstrate the flexible resolution which can be achieved with LOFAR, by imaging at 0.\sarc\ 6 and 1.\sarc\ 2 resolution, which are both complementary to the existing 6\sarc\ \ Deep Fields Data Release 1 \citep[DR1;][]{sabater_lofar_2021,tasse_lofar_2021}. This is another advantage over VLBI with instruments at GHz frequencies, which require either observing at different frequency bands or with other instruments to acquire information on the flux density of sources at other scales. As described in \S~5 of \citetalias{morabito_identifying_2022}, for sources which are unresolved in the 0.\sarc\ 3 resolution images and do not exhibit a radio excess over that predicted by the star formation rate (SFR) derived from SED fitting \citepalias[see][for more details]{best_lofar_2023}, we can separate the AGN activity from SF using a combination of the 0.\sarc\ 3 and 6\sarc\ \ resolution images.

The 2,483 sources in Lockman Hole \citep{sweijen_deep_2022} and now the 13,058 sources in ELAIS-N1 \citep{de_jong_into_2024}, represent a massive step towards robust statistical studies where the population can be broken into bins of redshift and stellar mass to investigate the AGN and SFG populations in detail. \cite{cochrane_lofar_2023} and \cite{kondapally_cosmic_2022} constructed radio luminosity functions (RLFs) from the DR1 and SED  for the star forming galaxies and radio-excess AGN, respectively. Here we construct RLFs, for the first time, by physical process rather than overall galaxy classification. %We do this by using a combination of the 0.\sarc\ 3 and 6\sarc\ \ resolution imaging from Lockman Hole and ELAIS-N1. 
We investigate the cosmic evolution of AGN activity and SF by dividing the sample into redshift bins out to $z = 2.5$.

In this paper, we first describe the data in \S~\ref{sec:data}, followed by an overview of the methods in \S~\ref{sec:methods}. Section~\ref{sec:results} describes the results, followed by 
% discussion and 
conclusions in 
%\S~\ref{sec:discussion} and
\S~\ref{sec:conclusions}
%, respectively
.

Throughout this paper, we use a $H_0 = 70$ km$\,$s$^{-1}\,$Mpc$^{-1}$, $\Omega_M = 0.3$, and $\Omega_{\Lambda}= 0.7$ cosmology. Radio flux density is defined as $S\propto \nu^{\alpha}$ where $\nu$ is frequency and $\alpha = -0.7$ is radio spectral index.

@@

\subsection{LoTSS Deep Fields DR1}
\label{subsec:dr1}
The LOFAR Two-metre Sky Survey (LoTSS) has both a wide-area \citep{shimwell_lofar_2017,shimwell_lofar_2021,shimwell_lofar_2022} and Deep Fields \citep{tasse_lofar_2021,sabater_lofar_2021} component. The Deep Fields Data Release 1 (DR1) included Boötes, Lockman Hole, and ELAIS-N1, reaching sensitivities of 30, 23, and 20 \muJybm , using $\sim$80, $\sim$100, and $\sim$160 hours of observations, respectively. The resolution of these images is all 6\sarc . Alongside these deep radio images and their associated catalogues, \cite{kondapally_lofar_2021} provided a careful compilation of the ancillary deep data at other wavebands (from far-infrared to ultraviolet) and cross-matching to the radio catalogues, followed by photometric redshifts and stellar mass estimates in \cite{duncan_lofar_2021}. Collectively these catalogues provide an excellent resource, and we refer the reader for more details to the individual papers cited above.

Using the multi-wavelength cross-matched catalogues, \citetalias{best_lofar_2023} carried out detailed SED fitting for all radio-detected sources across the three deep fields. Full details can be found in their paper, but here we outline the most relevant details. All sources detected above 5$\sigma$ in their respective radio images had their far-infrared (FIR) to ultraviolet (UV) data fit using four different SED fitting software packages, and a `consensus' value was found for the derived galaxy properties using information from the goodness-of-fits.
%These were: MAGPHYS \citep{da_cunha_simple_2008}, BAGPIPES \citep{carnall_inferring_2018,carnall_how_2019}, CIGALE \citep{burgarella_star_2005,noll_analysis_2009,boquien_cigale_2019}, and AGN\textsc{fitter} \citep{calistro_rivera_agnfitter_2016}. The first two (MAGPHYS and BAGPIPES) use an energy balance approach, where optical to UV energy absorbed by dust must be transferred to thermal emission in the FIR to sub-mm.  CIGALE also uses the same energy-balance approach, but includes an AGN component. The final code, AGN\textsc{fitter} is designed to work in cases where the UV light and IR emission are spatially distinct, and thus the energy balance assumption does not hold. Many AGN fall under this umbrella. Each galaxy was fit with all four codes (twice for CIGALE, using different AGN components) and a `consensus' value found for the derived galaxy properties using information from the goodness-of-fits.

The derived properties used in this paper are the consensus SFRs and the galaxy classifications which arise from the SED fitting. \citetalias{best_lofar_2023} use the empirical relationship between radio luminosity and SFR to define sources as either \textit{radio excess} or not. The AGN components used to fit the multi-wavelength information indicate whether a source is a radiatively efficient AGN (i.e., is there evidence for a disk/torus structure) or not. Combining these provides the following classes:
\begin{itemize}
    \item Radio excess + radiatively efficient (\textit{`radio-loud'})
    \item Radio excess + not radiatively efficient (\textit{`radio-loud'})
    \item Radiatively efficient + not radio excess (\textit{`radio-quiet'})
    \item Radiatively inefficient + not radio excess
\end{itemize}
The first three classes are AGN, and the final class is star forming galaxies (SFGs).  Five percent of all sources are unclassified, as they do not have a radio excess and the fits could not determine between radiatively efficient or inefficient.

@@

Data for LoTSS are recorded with all available LOFAR stations, which are spread across eight different European countries, with a preponderance of stations in the Netherlands. The longest baseline (currently Ireland to Poland) is $\sim$2,000 km, providing unprecedented resolution at MHz frequencies. While LoTSS uses only the stations in the Netherlands for its standard data processing, post-processing of LoTSS data to achieve sub-arcsecond resolution is ongoing both for individual sources with $S>10\,$mJy in the wide-area survey \citep{morabito_sub-arcsecond_2022} and for the entire field of view in the Deep Fields \citep[][Escott et al., in prep; Bondi et al., in prep]{sweijen_deep_2022,de_jong_into_2024}. %While the full field of view imaging is currently too computationally expensive for the wide area, we are focusing on the Deep Fields. 
Due to the fact that the size of international stations is larger than that of stations in the Netherlands, the station beams have a smaller field of view. Coupled with bandwidth and time smearing, this limits the field of view effectively to a radius of 1.24 deg, which still provides 6.25 deg$^2$ of coverage\footnote{The imaged field of view is square with 2.5 deg per side.}.

Both Lockman Hole and ELAIS-N1 have been imaged at 0.\sarc\ 3 resolution \citep[][respectively]{sweijen_deep_2022,de_jong_into_2024}. The resulting images have median rms noise of 34 \muJybm\ (Lockman Hole) and 17 \muJybm\ (ELAIS-N1), although the noise is lowest in the centre and increases radially in both images.  %(and can vary per facet). 
Catalogues from these images were processed to group individual Gaussian islands into complete sources, and remove duplicates. The resulting catalogues contain 2,316 and 13,058 sources above 5$\sigma$, for Lockman Hole and ELAIS-N1 respectively. Although \cite{de_jong_into_2024} also imaged ELAIS-N1 at 0.\sarc\ 6 and 1.\sarc\ 2, for consistency with Lockman Hole we use only the 0.\sarc\ 3 and 6\sarc\ resolution images.

@@

Thanks to the added collecting area of the international stations, which increases the sensitivity of the array, the 0.\sarc\ 3 resolution image for ELAIS-N1 has a lower noise level than the Deep Fields DR1. To ensure consistently built samples, and make use of DR1 information, we limit the high-resolution catalogues. For both fields, 
%The derived galaxy properties from \citetalias{best_lofar_2023} (see \S~\ref{subsec:dr1}) were for LoTSS Deep Fields DR1, and thus we have to limit the source catalogues we wish to use. For both Lockman Hole and ELAIS-N1 catalogues derived from the 0.\sarc\ 3 resolution images, 
we first find the local rms in the companion 6\sarc\ \ resolution noise maps\footnote{\label{dr1}Available at \href{https://lofar-surveys.org/deepfields.html}{https://lofar-surveys.org/deepfields.html}} at the position of each source in the 0.\sarc\ 3 resolution catalogue. Using the peak brightness of the source in the catalogue from the 0.\sarc\ 3 resolution image, we calculate the peak-to-noise ratio and only select sources which would be detectable at the 5$\sigma$ limit in the 6\sarc\ resolution image, effectively ensuring all sources should have a match in DR1. We then cross-match based on radio position, to provide the radio information from the catalogues derived from both the 0.\sarc\ 3 resolution image and the 6\sarc\ \ resolution image. We use optical star masks\textsuperscript{\ref{dr1}} to remove sources with multi-wavelength information affected by bright stars. Finally, we trim the area down of the multi-wavelength catalogue to the 2.5 deg$\times$2.5 deg area which is covered by the 0.\sarc\ 3 resolution image. We are left with \variable{output/lockman_detectable.txt}sources in Lockman Hole, and %\variable{output/en1_detectable.txt}sources in ELAIS-N1. %We use the source name from the catalogue derived from the 6\sarc\ \ resolution to cross-match to the catalogue from \citetalias{best_lofar_2023}.

We determine whether sources are resolved or unresolved (relevant in the next section) in the same manner as \S~3.1 of \cite{shimwell_lofar_2019,shimwell_lofar_2022}, by finding an upper limit to the intrinsically unresolved sources. Theoretically, these should have equal peak brightness and total flux density, but LOFAR maps are impacted by smearing due to the ionosphere, and high resolution images have time and bandwidth smearing effects. Thus intrinsically unresolved sources can appear spatially extended. To find this upper limit, we take all sources which are fit by a single Guassian component ({\ttfamily S\_CODE = `S'}), and are smaller than a multiple of the beam size. For the Lockman Hole, we use a factor of 3, but for ELAIS-N1 we find that a factor of 4 provided a better sample. This may be because the ELAIS-N1 field uses multiple nights, which have varying ionospheric conditions \citep[see \S~2 in][]{de_jong_into_2024}. 
%This may be because half of the data for ELAIS-N1 had been pre-averaged to 2 sec integration times, which will increase the time smearing \citep[see \S~2 in][]{de_jong_into_2024}. 
Plotting the total flux density to peak brightness ratio of this sample as a function of the local signal to noise, we fit a Sigmoid function to the upper 99.9th percentile. Applying this cut to the entire catalogue separates sources into unresolved or resolved. Any source which does not have {\ttfamily S\_CODE = `S'} is also considered resolved.
%We fit a Sigmoid function to find the upper 99.9th percentile of sources considering their total flux density to peak brightness ratio as a function of the local signal to noise. Anything above this boundary is considered to be resolved. We also treat sources which do not have {\ttfamily S\_CODE = `S'} as resolved.

@@

We follow the same method of constructing RLFs as in \cite{kondapally_cosmic_2022} and \cite{cochrane_lofar_2023}. The 0.\sarc\ 3 resolution image is smaller than the 6\sarc\ resolution image, so we first reproduce the published RLFs (Fig.~\ref{fig:rlfs}, left panel) using the galaxy classifications from \citetalias{best_lofar_2023} to check that we have implemented the method appropriately, and that the smaller area is consistent. We use the completeness corrections from the respective papers. The only correction we do not implement is the photo-$z$ correction used in \cite{cochrane_lofar_2023}. The authors note that this correction is small, and as it is not implemented in \cite{kondapally_cosmic_2022} we chose to keep the method for RLF construction in this paper consistent across both samples. We use the same redshift range, $0.003 < z < 0.3$, to verify that we have good agreement with the published RLFs.
